---
title: 'Experiment004: UMAP visualization of zircon'
author: "Joe Skufca"
date: "2023-04-20"
output:
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
  html_notebook:
    toc: yes
    toc_depth: 3
---

The primary object of this experiment is to visualize zircon characteristics using UMAP and (additionally) to embed those visualizations onto a mapview object via popup viewer. 


# Setup - Packages

```{r}
library(here)
library(readxl)
library(tidyverse)
library(gghighlight)
library(lubridate)
library(janitor)

library(fs)

library(rgl)
library(umap)
library(sf)
library(mapview)
library(maptools)
library(leafpop)
library(htmlwidgets)

```


# Load Data

Read the raw data:

```{r}
file1=here("data-raw","ME_Geochron_ALL_FILTERED_02.xlsx")
df_r=read_excel(file1,sheet=3) %>% clean_names() 

```

Extract a subset of the columns that are relevant.  We consider this as the "working" dataframe.

Some choices:

* Keep only one age calculation (207_206), which appears most accurate.
* Keep only information characterizing the sample, not the process (so - discard scientist name, equipment, date of measurement)
* ratio variables not retained as they could be recreated from raw variables if needed.

```{r}
df_w = df_r %>% rowid_to_column("zircon") %>%
  select(zircon,1:geological_domain,class:geochron_id,
                       final_age207_206:final_age207_206_prop2se,p:u_2se)

```

We create a dataframe focused specifically of the metric variables:

```{r message=FALSE, warning=FALSE}
df_m=df_w %>% select(-contains("2se")) %>% select(zircon:station_id,final_age207_206:u ) %>%
  mutate(across(final_age207_206:u,as.numeric))
```

We also construct a dataframe focused on meta-variables (and categorical variables) :

```{r}
df_meta=df_w %>% select(zircon:rock_type,age=final_age207_206)
```


## Data cleaning and prep

Visualize for missing data:

```{r fig.width=12}
naniar::vis_miss(df_m)
```

We eliminate the points for which we dont have multivariate data (beyond 1730).

```{r}
df_2 =df_m %>% slice_head(n=1730)
naniar::vis_miss(df_2)
```


#### Let's subset to columns with nearly complete data:

```{r}
df_v=df_2 %>% select(1:4,ti:sr,nb:pr,sm:gd,yb,hf,th,u)
naniar::vis_miss(df_v)
```

The missing date for phosphorus is BDL, which we will refill with 0s.

```{r message=FALSE, warning=FALSE, fig.width=12}

df_v2=df_v  %>% replace(is.na(.), 0) %>% slice_sample(prop=1) # subsample, if desired, to improve speed.

```

`df_v2` is our finalized dataset of zircon features.

# Umap coordinates and plotting 

We will use the zircon characteristics and the UMAP algorithm to construct a coordinate representation of the set of zircons. 



## Let's try this data by building UMAP coordinates.

It is our intention to color by age, we will use the other variables to construct the coordinate representation.

```{r, warning=FALSE, fig.width=12}
umap_data=df_v2 %>% select(-zircon,-station_id,-final_age207_206) %>%
  mutate(across(.fns=scales::rescale))

zirclist= df_v2 %>% select(zircon) %>% pull() # to use as id


umap_v=umap(umap_data,min_dist=0.0001,n_neighbors=30,metric="manhattan")

```

## Color by age

We will use a custom plotting function to construct both a faceted (where we may choose the variable) plot along with an overall plot.  Additionally, we output a dataframe which joints the UMAP coordinates with the metadata and categorical variables.

```{r plotfun2}
my_plot2=function(umap,zirc_list,var,facet=TRUE) {
  
  if (facet) {
  df1=as_tibble(umap$layout) %>%
    mutate(zircon=zirc_list) %>% 
    left_join(df_meta) 
  
    p=df1 %>% ggplot(aes(V1,V2,color=age)) +geom_point()+
    facet_wrap(vars({{var}}))+scale_color_viridis_c()+
    gghighlight()+theme_minimal()
  print(p)
  }
   p1=df1 %>% ggplot(aes(V1,V2,color=age)) +scale_color_viridis_c()+
     geom_point()+theme_minimal()
  print(p1)

  df1
}

```


```{r, warning=FALSE, fig.width=12}
df_out=my_plot2(umap_v,zirclist,var=geological_domain)
```


```{r fig.width=12, warning=FALSE}
df_out=my_plot2(umap_v,zirclist,var=station_id)
```

# Plotting on map

```{r}
df_sf=df_meta %>% group_by(station_id) %>% 
  select(station_id,latitude,longitude,zone,geological_domain,class,subclass,rock_type) %>% summarise_all(first) %>% ungroup() %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)


```


Now lets create individual plots for each rock:

```{r}
# custom plotting function
popplot=function(id){
  df_out %>% 
  filter(station_id==id) %>% 
  ggplot(aes(V1,V2,color=age))+
     geom_point(size=1)+theme_void()+scale_color_viridis_c()+ theme(legend.position = "none")+ theme(axis.text.x = element_text(size = 1),
    panel.background = element_rect(fill = NA),
    plot.background = element_rect(colour = "antiquewhite4")) +labs(x = NULL, y = NULL) -> p
  p
}

```

Create subplots for each rock where microchemical analysis is available.

```{r}
dfsf2=df_sf %>% semi_join(df_out, by="station_id") 

rocks=map(dfsf2$station_id,popplot) # list of ppoplots

```

Re-visualize
```{r }
mapviewOptions(fgb = FALSE, georaster = FALSE)
mapview(dfsf2,popup=popupGraph(rocks,width=120,height=100))

```

