---
title: "Experiment001: UMAP visualization of zircon"
author: "Joe Skufca"
date: "2023-02-15"
output: 
  html_notebook:
    toc: true
    toc_depth: 3
---

The primary object of this experiment is to visualize zircon characteristics using UMAP and to choose reasonable parameter values.



# Setup

```{r}
library(plotly)
library(here)
library(readxl)
library(tidyverse)
library(gghighlight)
library(lubridate)
library(janitor)

library(fs)

library(rgl)
library(umap)
```


# Load Data

Read the raw data:

```{r}
file1=here("data-raw","ME_Geochron_ALL_FILTERED_02.xlsx")
df_r=read_excel(file1,sheet=3) %>% clean_names() 

```

Extract a subset of the columns that are relevant.  We consider this as the "working" dataframe.

```{r}
df_w = df_r %>% rowid_to_column("zircon") %>%
  select(zircon,1:geological_domain,class:geochron_id,
                       final_age207_206:final_age207_206_prop2se,p:u_2se)
```

We also create a dataframe focuses specifically of the measure variables:

```{r message=FALSE, warning=FALSE}
df_m=df_w %>% select(-contains("2se")) %>% select(zircon:station_id,final_age207_206:u ) %>%
  mutate(across(final_age207_206:u,as.numeric))
```

Also, we pull a dataframe focused on meta-varialbes:

```{r}
df_meta=df_w %>% select(zircon:rock_type)
```


Visualize for missing data:

```{r fig.width=12}
naniar::vis_miss(df_m)
```

# A first visualization


We will start by using only the dataset of complete data:



```{r}
df_comp=df_m %>% drop_na()
```


To keep the problem small (for initial testing), we will use only a subsample of those points.

```{r}
df_100=df_comp %>% slice_sample(prop = .5)
```


Separate numeric data from label.

```{r}
umap_data=df_100 %>% select(-zircon,-station_id) %>%
  mutate(across(.fns=scale))
zirc_list_100=df_100 %>% select(zircon) %>% pull()
```




Fit the umap model.
```{r}
umap_100=umap(umap_data)
```

```{r}
my_plot=function(umap,zirc_list,var,facet=TRUE) {
  
  if (facet) {
  df1=as_tibble(umap$layout) %>%
    mutate(zircon=zirc_list) %>% 
    left_join(df_meta) 
  
    p=df1 %>% ggplot(aes(V1,V2,color={{var}})) +geom_point()+
    facet_wrap(vars({{var}}))+
    gghighlight()+theme_minimal()
  
  
  print(p)
  }
  
   p1=df1 %>% ggplot(aes(V1,V2,color={{var}})) +geom_point()+theme_minimal()
  
  
  print(p1)
  
  
  
  
  df1
}


df_out=my_plot(umap_100,zirc_list_100,var=station_id)
```


Fit the umap model.

Change mindist
```{r}
umap_100=umap(umap_data,min_dist=0.0001,n_neighbors=10,metric="manhattan")
df_out=my_plot(umap_100,zirc_list_100,var=station_id)
```




### Let's try using more data.


We elimate the points for which we dont have multivariate data (beyond 1730).

```{r}
df_2 =df_m %>% slice_head(n=1730)
naniar::vis_miss(df_2)
```


#### What if we complete the dataset with 0s.
```{r}
df_v=df_2 %>% replace(is.na(.), 0)
```


Let's revisualize, working up slowly to the full dataset:

```{r}
df_v2=df_v %>% slice_sample(prop=.8)


umap_data=df_v2 %>% select(-zircon,-station_id) %>%
  mutate(across(.fns=scales::rescale))
zirclist= df_v2 %>% select(zircon) %>% pull()


umap_v=umap(umap_data,min_dist=0.0001,n_neighbors=10,metric="manhattan")

df_out=my_plot(umap_v,zirclist,var=station_id)
```

#### Let's subset to columns with nearly complete data:

```{r}
df_v=df_2 %>% select(1:4,ti:sr,nb:pr,sm:gd,yb,hf,th,u)
naniar::vis_miss(df_v)
```


Let's revisualize, working up slowly to the full dataset:

```{r message=FALSE, warning=FALSE, fig.width=12}
df_v2=df_v  %>% drop_na() %>% slice_sample(prop=1)


umap_data=df_v2 %>% select(-zircon,-station_id) %>%
  mutate(across(.fns=scales::rescale))
zirclist= df_v2 %>% select(zircon) %>% pull()


umap_v=umap(umap_data,min_dist=0.0001,n_neighbors=30,metric="manhattan")

```
 
```{r, warning=FALSE, fig.width=12}
df_out=my_plot(umap_v,zirclist,var=station_id)
```

```{r, warning=FALSE, fig.width=12}
df_out=my_plot(umap_v,zirclist,var=zone)
```

```{r, warning=FALSE, fig.width=12}
df_out=my_plot(umap_v,zirclist,var=geological_domain)
```

```{r, warning=FALSE, fig.width=12}
df_out=my_plot(umap_v,zirclist,var=rock_type)
```

## Anything interesting in 3d vis?


```{r}
umap_v3=umap(umap_data,min_dist=0.0001,n_neighbors=30,metric="manhattan",n_components=3)
```

```{r fig.width=12}
# umap_labels=df_v2 %>% select(zircon) %>%
#   left_join(df_meta) %>%  pull(rock_type)

umap_labels=df_v2 %>% select(zircon) %>%
   left_join(df_meta)

 df3=as_tibble(umap_v3$layout) %>%
    bind_cols(umap_labels)
 
 #df1 %>% ggplot(aes(V1,V2,color=g)) +geom_point()+theme_minimal()
 
 df3 %>% ggplot(aes(V1,V2,color=geological_domain)) +geom_point()+theme_minimal()+
   facet_wrap(~station_id)
 
 
```

```{r}
plot_ly(df3, x = ~V1, y = ~V2, z = ~V3, color = ~geological_domain,size=3)

```

# Other data studies

Lets check for correlations:

```{r}
library(GGally)
pairs(umap_data %>% select(8:13))
```


What does PCA reveal?

```{r}
my_pca <- prcomp(umap_data)
summary(my_pca)
```

10 principal compents is sufficient.

```{r fig.width=12}
umap_p=umap(my_pca$x[,1:10],min_dist=0.0001,n_neighbors=30,metric="manhattan")

df_out=my_plot(umap_p,zirclist,var=station_id)
```

PCA seems to not be a good choice.

### Two variables


```{r fig.width=12}
  df1=as_tibble(umap_v$layout) %>%
    mutate(zircon=zirclist) %>% 
    left_join(df_meta) 
  
  df1 %>% ggplot(aes(V1,V2,color=class)) +geom_point()+
    facet_wrap(vars(zone,station_id))+
    theme_minimal()
  
```
