---
title: "Experiment002: UMAP visualization of zircon"
author: "Joe Skufca"
date: "2023-02-15"
output: 
  html_notebook:
    toc: true
    toc_depth: 3
---

The primary object of this experiment is to visualize zircon characteristics using UMAP and to choose reasonable parameter values.  

Experiment 3 coloring with age.


# Setup

```{r}
library(plotly)
library(here)
library(readxl)
library(tidyverse)
library(gghighlight)
library(lubridate)
library(janitor)

library(fs)

library(rgl)
library(umap)
```


# Load Data

Read the raw data:

```{r}
file1=here("data-raw","ME_Geochron_ALL_FILTERED_02.xlsx")
df_r=read_excel(file1,sheet=3) %>% clean_names() 

```

Extract a subset of the columns that are relevant.  We consider this as the "working" dataframe.

Some choices:

* Keep only one age calculation (207_206), which appears most accurate.
* Keep only information characterizing the sample, not the process (so - discard scientist name, equipment, date of measurement)
* ratio variables not retained as they could be recreated from raw variables if needed.

```{r}
df_w = df_r %>% rowid_to_column("zircon") %>%
  select(zircon,1:geological_domain,class:geochron_id,
                       final_age207_206:final_age207_206_prop2se,p:u_2se)



```

We also create a dataframe focused specifically of the metric variables:

```{r message=FALSE, warning=FALSE}
df_m=df_w %>% select(-contains("2se")) %>% select(zircon:station_id,final_age207_206:u ) %>%
  mutate(across(final_age207_206:u,as.numeric))
```

Also, we pull a dataframe focused on meta-variables (and categorical) :

```{r}
df_meta=df_w %>% select(zircon:rock_type,age=final_age207_206)
```


## Data cleaning and prep

Visualize for missing data:

```{r fig.width=12}
naniar::vis_miss(df_m)
```

We elimate the points for which we dont have multivariate data (beyond 1730).

```{r}
df_2 =df_m %>% slice_head(n=1730)
naniar::vis_miss(df_2)
```


#### Let's subset to columns with nearly complete data:

```{r}
df_v=df_2 %>% select(1:4,ti:sr,nb:pr,sm:gd,yb,hf,th,u)
naniar::vis_miss(df_v)
```

The missing date for phosphorus is BDL, which we will refill with 0s.

```{r message=FALSE, warning=FALSE, fig.width=12}
#df_v2=df_v  %>% drop_na() %>% slice_sample(prop=1)
df_v2=df_v  %>% replace(is.na(.), 0) %>% slice_sample(prop=1)

```

# Utility function for plotting

The helper function below will assist.

```{r plotfun}
my_plot=function(umap,zirc_list,var,facet=TRUE) {
  
  if (facet) {
  df1=as_tibble(umap$layout) %>%
    mutate(zircon=zirc_list) %>% 
    left_join(df_meta) 
  
    p=df1 %>% ggplot(aes(V1,V2,color={{var}})) +geom_point()+
    facet_wrap(vars({{var}}))+
    gghighlight()+theme_minimal()
  print(p)
  }
   p1=df1 %>% ggplot(aes(V1,V2,color={{var}})) +geom_point()+theme_minimal()
  print(p1)

  df1
}

```



### Let's try this data by building UMAP coordinates.


Let's revisualize, working up slowly to the full dataset:


 
```{r, warning=FALSE, fig.width=12}
umap_data=df_v2 %>% select(-zircon,-station_id) %>%
  mutate(across(.fns=scales::rescale))
zirclist= df_v2 %>% select(zircon) %>% pull()


umap_v=umap(umap_data,min_dist=0.0001,n_neighbors=30,metric="manhattan")

```
 
```{r, warning=FALSE, fig.width=12}
df_out=my_plot(umap_v,zirclist,var=station_id)
```

```{r, warning=FALSE, fig.width=12}
df_out=my_plot(umap_v,zirclist,var=zone)
```

```{r, warning=FALSE, fig.width=12}
df_out=my_plot(umap_v,zirclist,var=geological_domain)
```



# Color by age


```{r plotfun2}
my_plot2=function(umap,zirc_list,var,facet=TRUE) {
  
  if (facet) {
  df1=as_tibble(umap$layout) %>%
    mutate(zircon=zirc_list) %>% 
    left_join(df_meta) 
  
    p=df1 %>% ggplot(aes(V1,V2,color=age)) +geom_point()+
    facet_wrap(vars({{var}}))+scale_color_viridis_c()+
    gghighlight()+theme_minimal()
  print(p)
  }
   p1=df1 %>% ggplot(aes(V1,V2,color=age)) +scale_color_viridis_c()+
     geom_point()+theme_minimal()
  print(p1)

  df1
}

```

```{r, warning=FALSE, fig.width=12}
df_out=my_plot2(umap_v,zirclist,var=station_id)
```
